---
title: "Iterator objects"
author: "Dr Gavin Rhys Lloyd"
package: structToolbox
output: 
    html_document:
        keep_md: true
    github_document:
        df_print: kable
        html_preview: false
vignette: >
  %\VignetteIndexEntry{Iterator objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = 'center'
)
library(structToolbox)
library(gridExtra)
```

</br></br>

# Introduction
Validation is an important aspect of chemometric modelling. The `STRUCT` framework enables this kind of iterative model testing through `iterator` objects. In order to demonstrate this we will first load the iris data set, which as been pre-prepared as a dataset object as part of the `STRUCT` package.

```{r}
D = iris_dataset()
summary(D)
```

</br></br>

# Cross-validation
Cross validation is a common technique for assessing the performance of classification models. For this example we will use a PLSDA model. Data should be mean centred prior to PLS, so we will build a model sequence first.

```{r}
M = mean_centre() + PLSDA(number_components=2,factor_name='Species')
M
```

Iterators objects like the k-fold cross-validation object can be created just like any other struct object. Parameters can be set at creation =, and accessed/changed later using dollar notation.

```{r}
XCV = kfold_xval(folds=5,factor_name='Species')
# change the number of folds
XCV$folds=10
XCV$folds
```

The model to be cross-validated can be set/accessed used the `models` method.

```{r}
models(XCV)=M
models(XCV)
```

Alternatively,  iterators can be combined with models using the multiplication symbol:
```{r}
XCV = kfold_xval(folds=5,method='venetian',factor_name='Species') * 
      (mean_centre()+PLSDA(number_components = 2,factor_name='Species'))
```

The `run` method can be used with any iterator object. The iterator will then run the model sequence multiple times. In our case we will run cross-validation 5 times splitting the data into different training and test sets each time. The `run` method also needs a `metric` to be specified. This metric may be calculated once after all iterations, or after each iteration, depending on the iterator type (resampling, permutation etc). For cross-validation we will calculate balanced accuracy after all iterations.

```{r}
XCV = run(XCV,D,balanced_accuracy())
XCV$metric
```

</br></br>

Like other `STRUCT` objects, iterators can have chart objects associated with them. The `chart.names` function will list them for an object.

```{r}
chart.names(XCV)
```

Charts for iterator objects can be plotted in the same way as charts for any other object.

```{r,warning=FALSE}
C = kfoldxcv_grid()
chart.plot(C,XCV)
```

It is possible to combine multiple iterators by multiplying them together. This is equivalent to nesting one iterator inside the other. For example, we can repeat our cross-validation multiple times by permuting the sample order.

```{r}
P = permute_sample_order(number_of_permutations = 10) * 
    kfold_xval(folds=5,factor_name='Species')*
    (mean_centre() + PLSDA(factor_name='Species',number_components=2))
P = run(P,D,balanced_accuracy())
P$metric
```







