---
title: "Introduction to 'struct' objects"
author: 
- name: Gavin R Lloyd
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: g.r.lloyd@bham.ac.uk
- name: Andris Jankevics
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: a.jankevics@bham.ac.uk
- name: Ralf J Weber
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: r.j.weber@bham.ac.uk
date: "`r Sys.Date()`"
output: html_document
---


```{r, include=FALSE}
knitr::opts_chunk$set(
    dpi=96,fig.width=5,fig.height=5.5,fig.retina = 1,fig.small = TRUE
)
set.seed(57475)

suppressPackageStartupMessages({
    # Bioconductor packages
    library(structToolbox)
    library(pmp)
    library(ropls)
    library(BiocFileCache)
    library(BiocStyle)
  
    # CRAN libraries
    library(ggplot2)
    library(gridExtra)
    library(cowplot)
    library(openxlsx)
})


# use the BiocFileCache
bfc <- BiocFileCache(ask = FALSE)
```
# Introduction to `struct` objects, including models, model sequences, model charts and ontology.

## Introduction
PCA (Principal Component Analysis) and PLS (Partial Least Squares) are commonly applied methods for exploring and analysing multivariate datasets. Here we use these two statistical methods to demonstrate the different types of `struct` (STatistics in R Using Class Templates) objects that are available as part of the `structToolbox` and how these objects (i.e. class templates) can be used to conduct unsupervised and supervised multivariate statistical analysis.

## Dataset
For demonstration purposes we will use the "Iris" dataset. This famous (Fisher's or Anderson's) dataset contains measurements of sepal length and width and petal length and width, in centimeters, for 50 flowers from each of 3 class of Iris. The class are Iris setosa, versicolor, and virginica. See here (https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/iris.html) for more information.

Note: this vignette is also compatible with the Direct infusion mass spectrometry metabolomics "benchmark" dataset described in Kirwan et al., _Sci Data_ *1*, 140012 (2014) (https://doi.org/10.1038/sdata.2014.12).

Both datasets are available as part of the structToolbox package and already prepared as a `DatasetExperiment` object.

```{r}
## Iris dataset (comment if using MTBLS79 benchmark data)
D = iris_DatasetExperiment()
D$sample_meta$class = D$sample_meta$Species

## MTBLS (comment if using Iris data)
# D = MTBLS79_DatasetExperiment(filtered=TRUE)
# M = pqn_norm(qc_label='QC',factor_name='sample_type') + 
#   knn_impute(neighbours=5) +
#   glog_transform(qc_label='QC',factor_name='sample_type') +
#   filter_smeta(mode='exclude',levels='QC',factor_name='sample_type')
# M = model_apply(M,D)
# D = predicted(M)

# show info
D
```

### DatasetExperiment objects
The `DatasetExperiment` object is an extension of the `SummarizedExperiment` class used by the Bioconductor community. It contains three main parts:

1. `data` A data frame containing the measured data for each sample.
2. `sample_meta` A data frame of additional information related to the samples e.g. group labels.
3. `variable_meta` A data frame of additional information related to the variables (features) e.g. annotations

Like all `struct` objects it also contains `name` and `description` fields (called "slots" is R language). 

A key difference between `DatasetExperiment` and `SummarizedExperiment` objects is that the data is transposed. i.e. for `DatasetExperiment` objects the samples are in rows and the features are in columns, while the opposite is true for `SummarizedExperiment` objects.

All slots are accessible using dollar notation.

```{r}
# show some data
head(D$data[,1:4])
```

## Using `struct` model objects

### Statistical models
Before we can apply e.g. PCA we first need to create a PCA object. This object contains all the inputs, outputs and methods needed to apply PCA. We can set parameters such as the number of components when the PCA model is created, but we can also use dollar notation to change/view it later. 

```{r}
P = PCA(number_components=15)
P$number_components=5
P$number_components
```
  
The inputs for a model can be listed using `param_ids(object)`:

```{r}
param_ids(P)
```
Or a summary of the object can be printed to the console:
```{r}
P
```

### Model sequences
Unless you have good reason not to, it is usually sensible to mean centre the columns of the data before PCA. Using the `STRUCT` framework we can create a model sequence that will mean centre and then apply PCA to the mean centred data.

```{r}
M = mean_centre() + PCA(number_components = 4)
```
  
In `structToolbox` mean centring and PCA are both model objects, and joining them using "+" creates a model_sequence object. In a model_sequence the outputs of the first object (mean centring) are automatically passed to the inputs of the second object (PCA), which allows you chain together modelling steps in order to build a workflow.

The objects in the model_sequence can be accessed by indexing, and we can combine this with dollar notation. For example, the PCA object is the second object in our sequence and we can access the number of components as follows:

```{r}
M[2]$number_components
```

### Training/testing models
Model and model_sequence objects need to be trained using data in the form of a `DatasetExperiment` object. For example, the PCA model sequence we created (`M`) can be trained using the iris `DatasetExperiment` object ('D'). 

```{r}
M = model_train(M,D)
```
This model sequence has now mean centred the original data and calculated the PCA scores and loadings.
  
Model objects can be used to generate predictions for test datasets. For the PCA model sequence this involves mean centring the test data using the mean of training data, and the projecting the centred test data onto the PCA model using the loadings. The outputs are all stored in the model sequence and can be accessed using dollar notation. For this example we will just use the training data again (sometimes called autoprediction), which for PCA allows us to explore the training data in more detail.

```{r}
M = model_predict(M,D)
```



Sometimes models don't make use the training/test approach e.g. univariate statsitics, filtering etc. For these models the `model_apply` method can be used instead. For models that do provide training/test methods, `model_apply` applies autoprediction by default i.e. it is a short-cut for applying `model_train` and `model_predict` to the same data.

```{r}
M = model_apply(M,D)
```

The available outputs for an object can be listed and accessed like input params, using dollar notation:
  
```{r}
output_ids(M[2])
M[2]$scores
```

### Model charts
The `struct` framework includes chart objects. Charts associated with a model object can be listed.

```{r}
chart_names(M[2])
```

Like model objects, chart objects need to be created before they can be used. Here we will plot the PCA scores plot for our mean centred PCA model.

```{r,fig.width=5,fig.height=5}
C = pca_scores_plot(factor_name='class') # colour by class
chart_plot(C,M[2])
```
  
Note that indexing the PCA model is required because the `pca_scores_plot` object requires a PCA object as input, not a model_sequence.

If we make changes to the input parameters of the chart, `chart_plot` must be called again to see the effects.

```{r}
# add petal width to meta data of pca scores
M[2]$scores$sample_meta$example=D$data[,1]
# update plot
C$factor_name='example'
chart_plot(C,M[2])
```
  
The `chart_plot` method returns a ggplot object so that you can easily combine it with other plots using the `gridExtra` or `cowplot` packages for example.

```{r,fig.width=10,fig.small=FALSE}
# scores plot
C1 = pca_scores_plot(factor_name='class') # colour by class
g1 = chart_plot(C1,M[2])

# scree plot
C2 = pca_scree_plot()
g2 = chart_plot(C2,M[2])

# arange in grid
grid.arrange(grobs=list(g1,g2),nrow=1)
```

### Ontology
Within the `struct` framework (and `structToolbox`) an `ontology` slot is provided to allow for standardardised definitions for objects and its inputs and outputs using the Ontology Lookup Service (OLS).

For example, STATO is a general purpose STATistics Ontology (http://stato-ontology.org). From the webpage:

> Its aim is to provide coverage for processes such as statistical tests, their conditions of application, and information needed or resulting from statistical methods, such as probability distributions, variables, spread and variation metrics. STATO also covers aspects of experimental design and description of plots and graphical representations commonly used to provide visual cues of data distribution or layout and to assist review of the results.

The ontology for an object can be set by assigning the ontology term identifier to the ontology slot of any `struct_class` object at design time. The ids can be listed using `$` notation:

```{r}
# create an example PCA object
P=PCA()

# ontology for the PCA object
P$ontology
```

The `ontology` method can be used obtain more detailed ontology information. When `cache = NULL` the `struct` package will automatically attempt to use the OLS API (via the `rols` package) to obtain a name and description for the provided identifiers. Here we used cached versions of the ontology definitions provided in the `structToolbox` package to prevent issues connecting to the OLS API when building the package.

```{r}
ontology(P,cache = ontology_cache()) # set cache = NULL (default) for online use
```
Note that the `ontology` method returns definitions for the object (PCA) and the inputs/outputs (number_of_components).

## Validating supervised statistical models
Validation is an important aspect of chemometric modelling. The `struct` framework enables this kind of iterative model testing through `iterator` objects. 

### Cross-validation
Cross validation is a common technique for assessing the performance of classification models. For this example we will use a Partial least squares-discriminant analysis (PLS-DA) model. Data should be mean centred prior to PLS, so we will build a model sequence first.


```{r}
M = mean_centre() + PLSDA(number_components=2,factor_name='class')
M
```

`iterator` objects like the k-fold cross-validation object (`kfold_xval`) can be created just like any other `struct` object. Parameters can be set at creation using the equals sign, and accessed or changed later using dollar notation.

```{r}
# create object
XCV = kfold_xval(folds=5,factor_name='class')

# change the number of folds
XCV$folds=10
XCV$folds
```

The model to be cross-validated can be set/accessed using the `models` method.

```{r}
models(XCV)=M
models(XCV)
```

Alternatively, iterators can be combined with models using the multiplication symbol was shorthand for the `models` assignement method:

```{r}
# cross validation of a mean centred PLSDA model
XCV = kfold_xval(
        folds=5,
        method='venetian',
        factor_name='class') * 
      (mean_centre() + PLSDA(factor_name='class'))
```


The `run` method can be used with any `iterator` object. The iterator will then run the set model or model sequence multiple times. 

In this case we run cross-validation 5 times, splitting the data into different training and test sets each time. 

The `run` method also needs a `metric` to be specified, which is another type of `struct` object. This metric may be calculated once after all iterations, or after each iteration, depending on the iterator type (resampling, permutation etc). For cross-validation we will calculate "balanced accuracy" after all iterations.

```{r}
XCV = run(XCV,D,balanced_accuracy())
XCV$metric
```

Note The `balanced_accuracy` metric actually reports 1-accuracy, so a value of 0 indicates perfect performance. The standard deviation "sd" is NA in this example because there is only one permutation.

Like other `struct` objects, iterators can have chart objects associated with them. The `chart_names` function will list them for an object.

```{r}
chart_names(XCV)
```

Charts for `iterator` objects can be plotted in the same way as charts for any other object.

```{r,warning=FALSE,fig.width=8,fig.height=4,fig.small=FALSE}
C = kfoldxcv_grid(
  factor_name='class',
  level=levels(D$sample_meta$class)[2]) # first level
chart_plot(C,XCV)
```

It is possible to combine multiple iterators by using the multiplication symbol. This is equivalent to nesting one iterator inside the other. For example, we can repeat our cross-validation multiple times by permuting the sample order.


```{r}
# permute sample order 10 times and run cross-validation
P = permute_sample_order(number_of_permutations = 10) * 
    kfold_xval(folds=5,factor_name='class')*
    (mean_centre() + PLSDA(factor_name='class',number_components=2))
P = run(P,D,balanced_accuracy())
P$metric
```