---
title: "Partial Least Squares (PLS) analysis of a untargeted LC-MS-based clinical metabolomics dataset."
author: 
- name: Gavin R Lloyd
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: g.r.lloyd@bham.ac.uk
- name: Andris Jankevics
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: a.jankevics@bham.ac.uk
- name: Ralf J Weber
  affiliation: Phenome Centre Birmingham, University of Birmingham, UK
  email: r.j.weber@bham.ac.uk
date: "`r Sys.Date()`"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(
    dpi=96,fig.width=5,fig.height=5.5,fig.retina = 1,fig.small = TRUE
)
set.seed(57475)

suppressPackageStartupMessages({
    # Bioconductor packages
    library(structToolbox)
    library(pmp)
    library(ropls)
    library(BiocFileCache)
    library(BiocStyle)
  
    # CRAN libraries
    library(ggplot2)
    library(gridExtra)
    library(cowplot)
    library(openxlsx)
})


# use the BiocFileCache
bfc <- BiocFileCache(ask = FALSE)
```

## Introduction
The aim of this vignette is to demonstrate how to 1) apply and validate Partial Least Squares (PLS) analysis using the structToolbox, 2) reproduce statistical analysis in [Thevenot et al. (2015)](https://pubs.acs.org/doi/10.1021/acs.jproteome.5b00354) and 3. compare different implementations of PLS.

## Dataset
The objective of the original study was to:

>  ...study the influence of age, body mass index (bmi), and gender on metabolite concentrations in urine, by analysing 183 samples from a cohort of adults with liquid chromatography coupled to high-resolution mass spectrometry.

[Thevenot et al. (2015)](https://pubs.acs.org/doi/10.1021/acs.jproteome.5b00354)

The "Sacurine" dataset needs to be converted to a `DatasetExperiment` object. The `r Biocpkg("ropls")` package provides the data as a list containing a `dataMatrix`, `sampleMetadata` and `variableMetadata`.


```{r}
data('sacurine',package = 'ropls')
# the 'sacurine' list should now be available

# move the annotations to a new column and rename the features by index to avoid issues
# later when data.frames get transposed and names get checked/changed
sacurine$variableMetadata$annotation=rownames(sacurine$variableMetadata)
rownames(sacurine$variableMetadata)=1:nrow(sacurine$variableMetadata)
colnames(sacurine$dataMatrix)=1:ncol(sacurine$dataMatrix)

# create DatasetExperiment
DE = DatasetExperiment(data = data.frame(sacurine$dataMatrix),
                       sample_meta = sacurine$sampleMetadata,
                       variable_meta = sacurine$variableMetadata,
                       name = 'Sacurine data',
                       description = 'See ropls package documentation for details')

# print summary
DE
```

## Data preprocessing
The Sacurine dataset used within this vignette has already been pre-processed:

> After signal drift and batch effect correction of intensities, each urine profile was normalized to the osmolality of the sample. Finally, the data were log10 transformed.

## Exploratory data analysis
Since the data has already been processed the data can be visualised using Principal Component Analysis (PCA) without further pre-processing. The `ropls` package automatically applies unit variance scaling (autoscaling) by default. The same approach is applied here.

```{r,fig.width=15,fig.height=6.5,fig.wide = TRUE}
# prepare model sequence
M = autoscale() + PCA(number_components = 5)
# apply model sequence to dataset
M = model_apply(M,DE)

# pca scores plots
g=list()
for (k in colnames(DE$sample_meta)) {
    C = pca_scores_plot(factor_name = k)
    g[[k]] = chart_plot(C,M[2])
}
# plot using cowplot
plot_grid(plotlist=g, nrow=1, align='vh', labels=c('A','B','C'))
```

The third plot coloured by gender (C) is identical to Figure 2 of the `r Biocpkg("ropls")` package vignette. The `structToolbox` package provides a range of PCA-related diagnostic plots, including D-statistic, scree, and loadings plots. These plots can be used to further explore the variance of the data.

```{r,fig.height=10,fig.width=9,fig.wide=TRUE}
C = pca_scree_plot()
g1 = chart_plot(C,M[2])

C = pca_loadings_plot()
g2 = chart_plot(C,M[2])

C = pca_dstat_plot(alpha=0.95)
g3 = chart_plot(C,M[2])

p1=plot_grid(plotlist = list(g1,g2),align='h',nrow=1,axis='b')
p2=plot_grid(plotlist = list(g3),nrow=1)
plot_grid(p1,p2,nrow=2)
```

## Partial Least Squares (PLS) analysis
The `ropls` package uses its own implementation of the (O)PLS algorithms. `structToolbox` uses the `pls`
package, so it is interesting to compare the outputs from both approaches. For simplicity only the scores
plots are compared.

```{r, fig.width=5, fig.height=5}
# prepare model sequence
M = autoscale() + PLSDA(factor_name='gender')
M = model_apply(M,DE)

C = pls_scores_plot(factor_name = 'gender')
chart_plot(C,M[2])
```

The plot is similar to fig.3 of the `r Biocpkg("ropls")` vignette. Differences are due to inverted LV axes, a common occurrence with the NIPALS algorithm (used by both `structToolbox` and `ropls`) which depends on how the algorithm is initialised.

To compare the R2 values for the model in structToolbox we have to use a regression model, instead of a discriminant model. For this we convert the gender factor to a numeric variable before applying the model.

```{r,fig.width=10,fig.height=9,fig.wide=TRUE}
# convert gender to numeric
DE$sample_meta$gender=as.numeric(DE$sample_meta$gender)

# models sequence
M = autoscale(mode='both') + PLSR(factor_name='gender',number_components=3)
M = model_apply(M,DE)

# some diagnostic charts
C = plsr_cook_dist()
g1 = chart_plot(C,M[2])

C = plsr_prediction_plot()
g2 = chart_plot(C,M[2])

C = plsr_qq_plot()
g3 = chart_plot(C,M[2])

C = plsr_residual_hist()
g4 = chart_plot(C,M[2])

plot_grid(plotlist = list(g1,g2,g3,g4), nrow=2,align='vh')
```

The `ropls` package automatically applies cross-validation to asses the performance of the PLSDA model. In `structToolbox` this is applied separately to give more control over the approach used if desired. The default cross-validation used by the `ropls` package is 7-fold cross-validation and we replicate that here.

```{r,results='asis'}
# model sequence
M = kfold_xval(folds=7, factor_name='gender') * 
    (autoscale(mode='both') + PLSR(factor_name='gender'))
M = run(M,DE,r_squared())

```
```{r,results='asis',echo=FALSE}
# training set performance
cat('Training set R2:\n')
cat(M$metric.train)
cat('\n\n')
# test set performance
cat('Test set Q2:\n')
cat(M$metric.test)
```

The validity of the model can further be assessed using permutation testing. For this we will return to a discriminant model.

```{r,fig.width=5,fig.height=5}
# reset gender to original factor
DE$sample_meta$gender=sacurine$sampleMetadata$gender
# model sequence
M = permutation_test(number_of_permutations = 10, factor_name='gender') *
    kfold_xval(folds=7,factor_name='gender') *
    (autoscale() + PLSDA(factor_name='gender',number_components = 3))
M = run(M,DE,balanced_accuracy())

C = permutation_test_plot(style='boxplot')
chart_plot(C,M)+ylab('1 - balanced accuracy')
```

The permuted models have a balanced accuracy of around 50%, which is to be expected for a dataset with two groups. The unpermuted models have a balanced accuracy of around 90% and is therefore much better than might be expected to occur by chance.